---
title: "Class 6: R Functions"
author: "Qingyun Zheng (PID:A16338254)"
format: pdf
toc: true
---

All functions in R have at least 3 things:

- A **name**, we pick this and use it to call our function,
- Input **arguments** (there can be multiple)
- The **body** lines of R code that do the work

## Our first (silly) function

Write a function to add some numbers

```{r}
add <- function(x,y=1) {
  x+y
}
```
Note: The y=1 means that if we don't provide a value for y, it will default to 1.


Now we call this function:

```{r}
add(5,10)
add(5)
```

Input values as vectors
```{r}
add(c(10, 10, 100))
add(c(10, 10), 100)
```



## Our second function

Write a function to generate random nucleotide sequencesof a user specified length

The `sample()` function can be helpful here
```{r}
sample(c("A","T","G","C"), size = 50, replace = TRUE) 
```
*Replace = TRUE fixes the size issue

I want the a 1 element long character vector that looks have no gaps like "ATCGCTA"

```{r}
v <- sample(c("A","T","G","C"), size = 50, replace = TRUE) 
paste(v, collapse = "")
```
*The paste(x, collapse = "") function removes the gap


Turn the above into a function that takes input length and generates nucleotide sequence of that input length
```{r}
generate_DNA <- function(x){
  v <- sample(c("A","T","G","C"), size = x, replace = TRUE)
  paste(v, collapse = "")
}
```

Callling the function above
```{r}
generate_DNA(20)
```


The if statement has the following structure:
```{r}
if(TRUE){
  cat("HELLO you!")
} else{
  cat("No you don't")
}
```


Add the ability to return a multi-element vector or a single element fasta like vector

```{r}
generate_fasta <- function(size = 50, fasta = TRUE){
  v <- sample(c("A","T","G","C"), size = size, replace = TRUE)
  s <- paste(v, collapse = "")
  
if(fasta){
  return(s)
} else{
  return(v)
}
}
```

```{r}
generate_fasta(10, fasta = FALSE)
generate_fasta()
generate_fasta(10)
```


## A protein generating function

```{r}
generate_protein <- function(size = 50, fasta = TRUE){
  aa <- c("A","R","N","D","C","Q","E","G","H","I",
          "L","K","M","F","P","S","T","W","Y","V")
  v <- sample(aa, size = size, replace = TRUE)
  s <- paste(v, collapse = "")
  
  if(fasta){
    return(s)
  } else{
    return(v)
  }
}
```

```{r}
generate_protein(60)
```


Use our new `generate_protein()` function to make random protein sequences of length 6 to 12 (i.e. one length = 6, one length = 7...)

Using `for` loop
```{r}
lengths <- 6:12
for (i in lengths){
  cat(">",i, sep = "")
  cat("\n")
  aa <- generate_protein(i)
  cat(aa)
  cat("\n")
}
```

```{r}

sapply(6:10, generate_protein)
```





